Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-14 talk/pepm2011/paper.tex
Typos.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

@@

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
objects that are allocated in the trace \emph{static} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of various Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco 1.6
\cite{rigo_representation-based_2004}, an extension to CPython which is a
just-in-time compiler that produces machine code at run-time. It is not based
on traces. Finally, we used two versions of PyPy's Python interpreter (revision
77823 of SVN trunk\footnote{\texttt{https://codespeak.net/svn/pypy/trunk}}): one
including the JIT but not optimizing the traces, and one using the allocation
removal optimizations (as well as some minor other optimizations, such as
constant folding).

@@

\begin{figure*}
\begin{center}
\begin{tabular}{|l||r|r|r|r|}
\hline
	&Cpython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
\hline
crypto\_pyaes	&2757.80 $\pm$ 0.98	&67.90 $\pm$ 0.47	&1652.00 $\pm$ 4.00 &266.86 $\pm$ 5.94 \\
django	&993.19 $\pm$ 0.50	&913.51 $\pm$ 4.22	&694.73 $\pm$ 2.86 &259.53 $\pm$ 1.79 \\
fannkuch	&1987.22 $\pm$ 2.02	&944.44 $\pm$ 0.61	&566.99 $\pm$ 1.06 &466.87 $\pm$ 1.85 \\
go	&947.21 $\pm$ 1.58	&445.96 $\pm$ 0.68	&2197.71 $\pm$ 25.21 &316.15 $\pm$ 9.33 \\
html5lib	&13987.12 $\pm$ 19.51	&17398.25 $\pm$ 36.50	&27194.45 $\pm$ 46.62	&10092.19 $\pm$ 23.50 \\
meteor-contest	&346.98 $\pm$ 0.35	&215.66 $\pm$ 0.23	&433.04 $\pm$ 1.45 &392.85 $\pm$ 0.87 \\
nbody\_modified	&637.90 $\pm$ 1.82	&256.78 $\pm$ 0.18	&135.55 $\pm$ 0.33 &103.93 $\pm$ 0.25 \\
pyflate-fast	&3169.35 $\pm$ 1.89	&1278.16 $\pm$ 3.13	&3285.89 $\pm$ 8.51	&1822.36 $\pm$ 11.52 \\
raytrace-simple	&2744.60 $\pm$ 51.72	&1072.66 $\pm$ 1.08	&2778.27 $\pm$ 15.13	&647.24 $\pm$ 5.44 \\
richards	&354.06 $\pm$ 1.00	&63.48 $\pm$ 0.15	&383.93 $\pm$ 3.28 &88.32 $\pm$ 0.91 \\
spambayes	&299.16 $\pm$ 0.35	&338.68 $\pm$ 3.14	&580.90 $\pm$ 24.68 &397.37 $\pm$ 10.60 \\
spectral-norm	&478.63 $\pm$ 0.80	&139.83 $\pm$ 1.54	&353.51 $\pm$ 1.39 &112.10 $\pm$ 1.17 \\
telco	&1207.67 $\pm$ 2.03	&730.00 $\pm$ 2.66	&1296.08 $\pm$ 4.37 &495.23 $\pm$ 2.14 \\
twisted\_names	&9.58 $\pm$ 0.01	&10.43 $\pm$ 0.01	&17.99 $\pm$ 0.27	&7.13 $\pm$ 0.09 \\
\hline
\end{tabular}
\end{center}
\caption{Benchmark Times in Milliseconds}
\label{fig:times}
\end{figure*}

>>>>

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

@@

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}.

@@

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of various Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against
Psyco\cite{rigo_representation-based_2004} 1.6,
an extension to CPython which is a
just-in-time compiler that produces machine code at run-time. It is not based
on traces. Finally, we used two versions of PyPy's Python interpreter (revision
77823 of SVN trunk\footnote{\texttt{http://codespeak.net/svn/pypy/trunk}}): one
including the JIT but not optimizing the traces, and one using the allocation
removal optimizations (as well as some minor other optimizations, such as
constant folding).

@@

\begin{figure*}
\begin{center}
\begin{tabular}{|l||r|r|r|r|}
\hline
	&CPython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
\hline
crypto\_pyaes	&2757.80 $\pm$ 0.98	&67.90 $\pm$ 0.47	&1652.00 $\pm$ 4.00 &266.86 $\pm$ 5.94 \\
django	&993.19 $\pm$ 0.50	&913.51 $\pm$ 4.22	&694.73 $\pm$ 2.86 &259.53 $\pm$ 1.79 \\
fannkuch	&1987.22 $\pm$ 2.02	&944.44 $\pm$ 0.61	&566.99 $\pm$ 1.06 &466.87 $\pm$ 1.85 \\
go	&947.21 $\pm$ 1.58	&445.96 $\pm$ 0.68	&2197.71 $\pm$ 25.21 &316.15 $\pm$ 9.33 \\
html5lib	&13987.12 $\pm$ 19.51	&17398.25 $\pm$ 36.50	&27194.45 $\pm$ 46.62	&10092.19 $\pm$ 23.50 \\
meteor-contest	&346.98 $\pm$ 0.35	&215.66 $\pm$ 0.23	&433.04 $\pm$ 1.45 &392.85 $\pm$ 0.87 \\
nbody\_modified	&637.90 $\pm$ 1.82	&256.78 $\pm$ 0.18	&135.55 $\pm$ 0.33 &103.93 $\pm$ 0.25 \\
pyflate-fast	&3169.35 $\pm$ 1.89	&1278.16 $\pm$ 3.13	&3285.89 $\pm$ 8.51	&1822.36 $\pm$ 11.52 \\
raytrace-simple	&2744.60 $\pm$ 51.72	&1072.66 $\pm$ 1.08	&2778.27 $\pm$ 15.13	&647.24 $\pm$ 5.44 \\
richards	&354.06 $\pm$ 1.00	&63.48 $\pm$ 0.15	&383.93 $\pm$ 3.28 &88.32 $\pm$ 0.91 \\
spambayes	&299.16 $\pm$ 0.35	&338.68 $\pm$ 3.14	&580.90 $\pm$ 24.68 &397.37 $\pm$ 10.60 \\
spectral-norm	&478.63 $\pm$ 0.80	&139.83 $\pm$ 1.54	&353.51 $\pm$ 1.39 &112.10 $\pm$ 1.17 \\
telco	&1207.67 $\pm$ 2.03	&730.00 $\pm$ 2.66	&1296.08 $\pm$ 4.37 &495.23 $\pm$ 2.14 \\
twisted\_names	&9.58 $\pm$ 0.01	&10.43 $\pm$ 0.01	&17.99 $\pm$ 0.27	&7.13 $\pm$ 0.09 \\
\hline
\end{tabular}
\end{center}
\caption{Benchmark Times in Milliseconds}
\label{fig:times}
\end{figure*}
